Network Data Model and BerlinMOD Benchmark

### 2010-07-09 Algebras/Network/BerlinMODNetwork/paper/PaperBerlinMODAndNetworkDataModel.tex
Implementation completed.

\documentclass[a4paper]{article}
\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{float}
\usepackage{longtable}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{todonotes}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data
models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two
representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data 
models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis
,HERMESMDCPelekis}
for DMFS and 
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,
NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

In NET the data type \dt{network} is modeled by two main components. One is the
set  of routes (streets) and the other one the set of junctions (crossings). The
domain of routes is defined as
\[
\begin{split}
Route = \{(id,\ l,\ c,\ kind,\ start)\ &|\ id \in \dt{int},\ l \in \dt{real},
\ c \in \dt{line},\\
&\ kind \in \{simple,\ dual\},\ start \in \{smaller,\ larger\}\},
\end{split}
\]
where $id$ is a distinct route idenitfier, $l$ is the length of the route, $c$
is the route curve as \dt{line} value (see Section \ref{sec:bmodbdatamod}),
$kind$ indicates if the lanes of the route are separated, and $start$ indicates
how the route curve is embedded into space.

@@

A single network position in a network $N_i$ of a set of networks 
$N=\{N_1,\ldots,N_k\}$
is represented by the data type \dt{gpoint}. The carrier set of \dt{gpoint} is 
defined as
\[\{(i,\ gp)\ |\ 1 \leq i \leq k \wedge gp \in RLoc(R) \cup \{ \perp \}\},\]
where $\perp$ again represents a undefined value.

A $route\ interval$ in $N$ is a pair of network positions on the same route. It
is represented by a quadruple $(rid,\ d_1,\ d_2,\ side)$, with
$(rid,\ d_1,\ side),\ (rid,\ d_2,\ side) \in Loc(N)$
and $d_1 \leq d_2$. Semantically
a $route interval$ represents all route locations $(rid,\ d,\ side)$ with
$d_1 \leq d \leq d_2$. A finite set of disjoint $route\ intervals$ of a network
$N$ is called region of $N$. The set of all possible regions in a network $N$ is
denoted as $Reg(N)$.

@@

The three relations have the following schemas:
\begin{ttfamily}
\begin{itemize}
  \item [] routes (id:int; length: real; curve: line; kind: bool; start: bool)
  \item [] junctions (r1id: int; r1rc: int; pos1: real; r2id: int; r2rc: int;\\
pos2: real; cc: int; pos: point)
  \item [] sections (rid: int; rrc: int; pos1: real; pos2: real; dual: bool;\\
length: real; curve: line)
\end{itemize}
\end{ttfamily}
As you can see, the $routes$ relation is equivalent to the domain of routes $Route$.
The tuple of the $junctions$ relation is somewhat different from $Junctions(R)$.
The record identifiers $r1rc$ and $r2rc$ support faster access to the corresponding
tuples in the $routes$ relation and the \dt{point} value $pos$ supports the 
connection
to the two dimensional plane.

@@

For the data types \dt{gpoint} and \dt{gline} \cite{NetworkGueting} proposes
the following implementations:
\begin{ttfamily}
\begin{itemize}
  \item [] gpoint: record \{nid: int; rid: int; pos: real; side: \{up,
down, none\};\}
  \item [] gline: record\{nid: int; rints: DBArray of record \{ rid: int;\\
pos1: real; pos2: real; side: \{up, down, none\};\}
\end{itemize}
\end{ttfamily}
For the data type \dt{mgpoint} the implementation consists of a set of \dt{ugpoint}.
The set is stored in a DBArray in ascending order of the time intervals of the
\dt{ugpoint}. Each \dt{ugpoint} is defined as:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{nid: int; rid: int; pos1: real; pos2: real;\\
side: \{up, down, none\}; t1: Instant; t2: Instant;\}
\end{itemize}
\end{ttfamily}
The \dt{ugpoint} is expected to move from $pos1$ to $pos2$ with constant speed on
route $rid$ in the given network $nid$ within the time interval defined by $t1$
and $t2$. Every time a \dt{mgpoint} changes the speed or changes the route a new
\dt{ugpoint} is written.

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section
\ref{sec:queries}\todo[caption={check reference}]{Check reference when paper
finished. Parts of query explanations might be moved to the appendix.}
together with our NDM algorithms for these queries.
\section{Implementation Issues}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

\label{sec:sortedgline}
The record of \dt{gline} was extended by an attribute $length$ of
\dt{real}, storing the length of the \dt{gline}, and an sorted flag, indicating
if the $route\ intervals$ in the DBArray are stored sorted or not. We call a
set of $route\ intervals$ sorted if it fullfills the following conditions:
\begin{itemize}
   \item all $route\ intervals$ are disjoint
   \item the $route\ intervals$ are stored in ascending order of their route identifiers
   \item if two disjoint $route\ intervals$ have the same route identifier, the
$route\ interval$ with the smaller start position is stored first
   \item for all $route\ intervals$, $start\ position \le end\ position$
\end{itemize}
We introduced this definition and sorted flag, because many algorithms take profit
from sorted \dt{gline} values. For example: If $n$ is the number of the
$route\ intervals$ in a \dt{gline}, the decision, if a \dt{gpoint} is inside
the \dt{gline} needs O($n$) time for unsorted and O($\log n$) time for sorted
\dt{gline} values.

@@

Sorting $route\ intervals$ takes time, when a \dt{gline} value is created. But
we think that this time is well invested, because the sorted \dt{gline} is computed
once, but supports many times in different algorithms a faster execution.

The implementation of the \dt{ugpoint} has been changed to:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{gp1: gpoint; gp2: gpoint; ti : Interval;\}
\end{itemize}
\end{ttfamily}
Where $ti$ is a time interval consisting of two time instants $t_1$ and $t_2$
and
two Boolean flags, indicating if $t_1$ respectively $t_2$ is part of the time
interval
or not.

At the same time the implementation of \dt{mgpoint} has been extended to:
\begin{ttfamily}
\begin{itemize}
  \item [] mgpoint: record \{units: DBArray of ugpoint; drivenDist: real;\\
trajectory: DBArray of sorted route intervals; trajDefined: bool;\\
mbr: rectangle3D\}
\end{itemize}
\end{ttfamily}
The DBArray of \dt{ugpoint} is consistent with the data model of \cite{NetworkGueting}.
The $drivenDist$ is the total length of all \dt{ugpoint} in the \dt{mgpoint}.
The DBArray of $route\ intervals$ represents all network positions ever traversed
by the \dt{mgpoint}. The flag indicates if the $trajectory$ is well defined,
because this attribute is not maintenanced in every operation changing \dt{mgpoint}
values. And the minimum spatio-temporal bounding box $mbr$ can be used for a preselection
in spatio-temporal queries.

@@

The spatio-temporal minimum bounding box was introduced as an attribute to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a
route may lead uphill in serpentine. A spatial bounding box only computed from
the spatial start and end position may not enclose all spatial positions of the
car within the unit. Therefore we always have to examine the spatial dimensions
of the $route\ interval$ passed within a unit to compute the units bounding box
using algorithm \ref{alg:unitbbox}.
\begin{algorithm}[H]
  \caption{\op{Berechnung Unit Bounding Box}}
  \label{alg:unitbbox}
  \begin{algorithmic}[1]
    \STATE Use B-Tree index of $routes$ relation to get route curve
    \STATE Extract subline of unit from route curve
    \STATE Compute bounding box of subline
    \STATE Add third dimension from unit time interval
  \end{algorithmic}
\end{algorithm}
If the number of routes in the $routes$ relation is $r$ the first step has a
time complexity of O($\log r$) time. If the route curve $r_i$ consists of $h_i$
line segments the time complexity of step 2 and 3 is O($h_i$) in the worst case.
Step 4 is done in O(1) time. Together we get a time complexity of
O($\log {r} + h_i$) to compute the bounding box of a single unit.

To compute the $mbr$ this computation must be done for each of the $m$ units of
the \dt{mgpoint} value. That takes O($m \log {r} + \sum_{i=1}^{m}{h_i}$) time.

If the $trajectory$ is defined we can reduce the time complexity of $mbr$
computation. The spatial extend of the $mbr$ can be computed
using the $t\ route intervals$ of the $trajectory$ similar to the $m$ units of
the \dt{mgpoint}. The third dimension will be added using the start and the end
time instant of the \dt{mgpoint} value. This algorithm has a time complexity of
O($t \log {r} + \sum_{i=1}^{t}{h_i}$), with $t \ll m$ in nearly all cases.

@@

We introduced Network Bounding Boxes (NBB) and Network Temporal Bounding Boxes
(NTBB) in our implementation to support indexing in terms of network positions
with R-Trees instead of spatial indexing.

Let \dt{ugpoint} and $route\ interval$ be defined like in \ref{sec:netdatamod}.
The NBB of a $route interval$ is defined as $(rid,\ rid,\ pos1,\ pos2)$,
which can be seen as an degenerated two dimensional rectangle.
The NTBB of a \dt{ugpoint} value is defined as $(rid,\ rid,\ pos1,\ pos2,\ t1,\ t2)$,
which analogous can be seen as an degenerated three dimensional rectangle.

We use these NBB and NTBB to create Network Position Indexes (NPI) and
Temporal Network Position Indexes (TNPI) indexing the network positions of
the \dt{mgpoint} values of the \bmodb{}. An NPI is an R-Tree build from the NBB
of the $route intervals$ of the $trajectory$ attributes of the \dt{mgpoint} values.
An TNPI is an R-Tree build from the NTBB of the \dt{ugpoint} values of the
\dt{mgpoint} values.
\subsection{Network Operations used in \bmodb{}}
\label{sec:implNetOperations}
In the sequel we give an overview of the operations on network objects used in
the network version of the \bmodb{}. The signatures of the simple operators are
shown in Table \ref{tab:simplenetoperations} together with a short explanation and
their time complexity.

We use $m$ (respectively $m_1,\ m_2$) as the number of units of the
\dt{mgpoint} value (resp. first, second \dt{mgpoint} value),
$r$ (resp. $r_1$, $r_2$) as the number of $route\ intervals$ of the \dt{gline}
value or the $trajectory$ attribute of an \dt{mgpoint} value (resp. of the
first, second argument), $R$ as the number of routes in the network, and $p$ as
the number of time intervals in a \dt{periods} value.
\begin{table}
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|l|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
  \op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
  \op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant& O($\log m$)\\
  \hline
  \op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the argument.&O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$)\\
  \hline
  \op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
  \op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
  \op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{table}

The operations \op{thenetwork}, \op{mpoint2mgpoint}, \op{line2gline}.
\op{point2gpoint}, and \op{polygpoints} which are used to construct the network
object, and to translate spatial and spatio-temporal values into network values
are described in Section  \ref{sec:Translation} in more detail.

For the Euclidean Distance computation we retranslate our network values into
spatial (\op{gline2line}) respectively spatio-temporal (\op{mgpoint2mpoint})
values.

The operation \op{gline2line} uses the Algorithm \ref{alg:gline2line} to
translate a \dt{gline} value into an \dt{line} value. 
\begin{algorithm}
  \caption{\op{gline2line}($gl$)}
  \label{alg:gline2line}
  \begin{algorithmic}[1]
    \FOR {each $route\ interval$ of $gl$}
      \STATE Use B-Tree index of $routes$ relation to get route curve of the
$route\ interval$
      \STATE Perform binary search on line segments of route curve to find the
      start position of the $route\ interval$
      \STATE Copy line segments to result until end pos of $route\ interval$ is
reached.
    \ENDFOR
    \RETURN resulting line
  \end{algorithmic}
\end{algorithm}
The loop from line 1 to
line 5 is repeated $r$ times. The B-Tree search in line 2 take O($\log{R}$)
time. The binary search in line 3 takes O($\log l_i$) time if the number of line
segments of route curve $Rr_i$ is $l_i$. The copy operation in line 4 takes
O($l_i$) time in the worst case. The return of the result has a worst case
time complexity of O($\sum_{i=1}^{r}{l_i}$). For the whole algorithm we
get
\[O(r \log{R} + \sum_{i=1}^{r}{\log{l_i}} + 2\sum_{i=1}^{r}{l_i}) 
= O(r \log{R} + \sum_{i=1}^{r}{l_i})\].

>>>>

\documentclass[a4paper]{article}
%\usepackage[left=3cm, right=3cm]{geometry}
\usepackage[utf8x]{inputenc}
\usepackage{float}
\usepackage{longtable}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{url}
\usepackage{array}
\usepackage{algorithmic}
\usepackage[ruled]{algorithm}
\usepackage{color}
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{caption}
\usepackage{todonotes}
\usepackage{rotating}
\pagestyle{headings}
\newcommand{\secondo}{\textsc{Secondo}}
\newcommand{\bmodb} {BerlinMOD Benchmark}
\newcommand{\op}[1]{\textbf{#1}}
\newcommand{\dt}[1]{\textsl{\underline{#1}}}
\newcommand{\true}{\textsl{TRUE}}
\newcommand{\false}{\textsl{FALSE}}
%opening
\title{Network Data Model and BerlinMOD Benchmark}
\author{Simone Jandt}
\date{Last Update: \today}
\begin{document}
\maketitle
\begin{abstract}
In the past, several data models for the representation of histories of
spatio-temporal data objects have been developed. We can categorize these data
models
into data models for objects moving freely in the two dimensional space and data
models for network constrained moving objects. In this paper we select two
representatives,
one for each data model category, which are both implemented in the \secondo{} DBMS,
and compare their capabilities with the \bmodb{}. We describe our implementation
of the used network constrained data model, the translation
from the \bmodb{} into the network constrained data model, and show
that in our experiments the network constrained data model outperforms the data
model of free movement in the two dimensional space by orders of magnitude.
\end{abstract}
\listoftodos
\section{Introduction}
In the past, several data models for the representation of
spatio-temporal data objects have been developed. We can categorize them into
data models for objects moving freely in two dimensional space (DMFS) and data
models
for network constrained moving objects (NCDM). For both categories several different
data models have been presented like
\cite{DataModelDataStructureGueting,RepresentingMovingObjectsGueting,STAUPelekis
,HERMESMDCPelekis}
for DMFS and
\cite{DynamicTransportNetworkDing,NetworkGueting,NetworkJensen,
NetworkVazirgiannis}
for NCDM, to name just a few. Objects which are restricted to use existing networks,
like cars are restricted to use road networks, can be represented as moving point
objects in both data models, whereas objects, which are not restricted by a given
network, like people, can be represented as moving point objects only in DMFS.

@@

In NET the data type \dt{network} is modeled by two main components. One is the
set  of routes (streets) and the other one the set of junctions (crossings). The
domain of routes is defined as
\[
\begin{split}
Route = \{(id,\ l,\ c,\ kind,\ start)\ &|\ id \in \dt{int},\ l \in \dt{real},
\ c \in \dt{line},\\
&\ kind \in \{simple,\ dual\},\\
&\ start \in \{smaller,\ larger\}\},
\end{split}
\]
where $id$ is a distinct route idenitfier, $l$ is the length of the route, $c$
is the route curve as \dt{line} value (see Section \ref{sec:bmodbdatamod}),
$kind$ indicates if the lanes of the route are separated, and $start$ indicates
how the route curve is embedded into space.

@@

Let $N=\{N_1,\ldots,N_k\}$ be a set of networks. A single network position in a
network $N_i$ is represented by the data type \dt{gpoint}. The carrier set of
\dt{gpoint} is defined as
\[\{(i,\ gp)\ |\ 1 \leq i \leq k \wedge gp \in RLoc(R) \cup \{ \perp \}\},\]
where $\perp$ again represents a undefined value.

A $route\ interval$ in $N_i$ is a pair of network positions on the same route. The $route\ interval$ is represented by a quadruple $(rid,\ d_1,\ d_2,\ side)$, with
$(rid,\ d_1,\ side),\ (rid,\ d_2,\ side) \in Loc(N)$
and $d_1 \leq d_2$. Semantically
a $route interval$ represents all route locations $(rid,\ d,\ side)$ with
$d_1 \leq d \leq d_2$. A finite set of disjoint $route\ intervals$ of a network
$N$ is called region of $N$. The set of all possible regions in a network $N$ is
denoted as $Reg(N)$.

@@

The three relations have the following schemas:
\begin{ttfamily}
\begin{itemize}
  \item [] routes (id:int; length: real; curve: line; kind: bool;\\
 start: bool)
  \item [] junctions (r1id: int; r1rc: int; pos1: real; r2id: int;\\
r2rc: int; pos2: real; cc: int; pos: point)
  \item [] sections (rid: int; rrc: int; pos1: real; pos2: real;\\
dual: bool; length: real; curve: line)
\end{itemize}
\end{ttfamily}
As you can see, the $routes$ relation is equivalent to the domain of routes $Route$.
The tuple of the $junctions$ relation is somewhat different from $Junctions(R)$.
The record identifiers $r1rc$ and $r2rc$ support faster access to the corresponding
tuples in the $routes$ relation and the \dt{point} value $pos$ supports the
connection
to the two dimensional plane.

@@

For the data types \dt{gpoint} and \dt{gline} \cite{NetworkGueting} proposes
the following implementations:
\begin{ttfamily}
\begin{itemize}
  \item [] gpoint: record \{nid: int; rid: int; pos: real;\\
side: \{up, down, none\};\}
  \item [] gline: record\{nid: int; rints: DBArray of record \{\\
rid: int; pos1: real; pos2: real;\\
side: \{up, down, none\};\}
\end{itemize}
\end{ttfamily}
For the data type \dt{mgpoint} the implementation consists of a set of \dt{ugpoint}.
The set is stored in a DBArray in ascending order of the time intervals of the
\dt{ugpoint}. Each \dt{ugpoint} is defined as:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{nid: int; rid: int; pos1: real;\\
pos2: real; side: \{up, down, none\};\\
t1: Instant; t2: Instant;\}
\end{itemize}
\end{ttfamily}
The \dt{ugpoint} is expected to move from $pos_1$ to $pos_2$ with constant speed on
route $rid$ in the given network $nid$ within the time interval defined by $t_1$
and $t_2$. Every time a \dt{mgpoint} changes the speed or changes the route a new
\dt{ugpoint} is written.

@@

The query set BerlinMOD/R includes 17 queries selected of the set of possible
combinations of the 5 aspects:
\begin{itemize}
  \item known or unknown object identity,
  \item standard attribute, spatial, temporal, or spatio-temporal dimension,
  \item point, range, or unbounded query interval,
  \item single object or object relationships condition type,
  \item with or without aggregation.
\end{itemize}
We will present the 17 queries in more detail in Section
\ref{sec:queries}\todo[caption={check reference}]{Check reference when paper
finished. Parts of query explanations might be moved to the appendix.}
together with our NDM algorithms for these queries.
\section{Implementation of NET}
\label{sec:implNDM}
We started a description of the implementation of NET in \secondo{} in Section
\ref{sec:netdatamod} with the description of the implementation proposed in
\cite{NetworkGueting}. In the meantime this implementation has been changed and
extended in some points to support faster query execution.

@@

\label{sec:sortedgline}
The record of \dt{gline} was extended by an attribute $length$ of
\dt{real}, storing the length of the \dt{gline}, and an sorted flag, indicating
if the $route\ intervals$ in the DBArray are stored sorted or not. We call a
set of $route\ intervals$ sorted if it fullfills the following conditions:
\begin{itemize}
   \item all $route\ intervals$ are disjoint
   \item the $route\ intervals$ are stored in ascending order of their route identifiers
   \item if two $route\ intervals$ have the same route identifier, the
$route\ interval$ with the smaller start position is stored first if the
$route\ intervals$ are disjoint
   \item for all $route\ intervals$, $start\ position \le end\ position$
\end{itemize}
We introduced this definition and sorted flag, because many algorithms take profit
from sorted \dt{gline} values. Let $r$ be the number of $route\ intervals$ in a
\dt{gline}, the decision, if a \dt{gpoint} is inside the \dt{gline} needs O($r$)
time for unsorted and O($\log r$) time for sorted \dt{gline} values.

@@

For sorting and compressing $route\ intervals$ we introduced a binary tree
data structure called $RITree$. This $RITree$ sorts and compresses the inserted
$route\ intervals$. If $r_i$ is the number of inserted $route\ intervals$ and
$r_{res}$ the number of resulting $route\ intervals$ sorting and compressing takes
O($r_i \log r_{res}$) time. The sorted $route\ intervals$ are returned in
O($r_{res}$) time. We think that this time is well invested, because the sorted
\dt{gline} is computed once, but many different algorithms can be executed faster
for sorted \dt{gline} values.

The implementation of the \dt{ugpoint} has been changed to:
\begin{ttfamily}
\begin{itemize}
  \item [] ugpoint: record \{gp1: gpoint; gp2: gpoint;\\
ti : Interval;\}
\end{itemize}
\end{ttfamily}
Where $t_i$ is a time interval consisting of two time instants $t_1$ and $t_2$
and
two Boolean flags, indicating if $t_1$ respectively $t_2$ is part of the time
interval
or not.

At the same time the implementation of \dt{mgpoint} has been extended to:
\begin{ttfamily}
\begin{itemize}
  \item [] mgpoint: record \{units: DBArray of ugpoint;\\
drivenDist: real; trajDefined: bool; mbr: rectangle3D\\
trajectory: DBArray of sorted $route\ intervals$;\}
\end{itemize}
\end{ttfamily}
The DBArray of \dt{ugpoint} is consistent with the data model of
\cite{NetworkGueting}.
The $drivenDist$ is the total length of all \dt{ugpoint} in the \dt{mgpoint}.
The DBArray of $route\ intervals$ represents all network positions ever traversed
by the \dt{mgpoint}. The flag indicates if the $trajectory$ is well defined,
because this attribute is not maintenanced in every operation changing
\dt{mgpoint} values. And the minimum spatio-temporal bounding box $mbr$ can be
used for a preselection in spatio-temporal queries.

@@

The spatio-temporal minimum bounding box was introduced as an attribute to the
\dt{mgpoint} because the computation of this value is very expensive in the
NDM. Although each unit of an \dt{mgpoint} stays on the same route
at the same speed it may follow different spatial directions. For example, a
route may lead uphill in serpentine. A spatial bounding box only computed from
the spatial start and end position may not enclose all spatial positions of the
car within the unit. Therefore we always have to examine the spatial dimensions
of the $route\ interval$ passed within a unit to compute the units bounding box
using algorithm \ref{alg:unitbbox}.
\begin{algorithm}[H]
  \caption{\op{Berechnung Unit Bounding Box}}
  \label{alg:unitbbox}
  \begin{algorithmic}[1]
    \STATE Get route curve usint B-Tree index of $routes$ relation
    \STATE Extract subline of unit from route curve
    \STATE Compute bounding box of subline
    \STATE Add third dimension from unit time interval
  \end{algorithmic}
\end{algorithm}
If the number of routes in the $routes$ relation is $R$ the first step has a
time complexity of O($\log R$) time. If the route curve $r_i$ consists of $l_i$
line segments the time complexity of step 2 and 3 is O($l_i$) in the worst case.
Step 4 is done in O(1) time. Together we get a time complexity of
O($\log {R} + l_i$) to compute the bounding box of a single unit.

To compute the $mbr$ this computation must be done for each of the $m$ units of
the \dt{mgpoint} value. That takes O($m \log {R} + \sum_{i=1}^{m}{l_i}$) time.

The time complexity can be reduced if the $trajectory$ is defined. We can use
the $r$ $route\ intervals$ of the $trajectory$ similar to the $m$ units of
the \dt{mgpoint} to compute the spatial bounding box. The third dimension can be
added using the start and the end time instant of the \dt{mgpoint} value. This a
algorithm has a time complexity of O($r \log {R} + \sum_{i=1}^{r}{l_i}$), with
$r \ll m$ in nearly all cases.

@@

We introduced Network Bounding Boxes (NBB) and Temporal Network Bounding Boxes
(TNBB) in our implementation to support indexing in terms of network positions
with R-Trees instead of spatial indexing.

Let \dt{ugpoint} and $route\ interval$ be defined like in \ref{sec:netdatamod}.
The NBB of a $route interval$ is defined as $(rid,\ rid,\ pos1,\ pos2)$,
which can be seen as an degenerated two dimensional rectangle.
The TNBB of a \dt{ugpoint} value is defined as
$(rid,\ rid,\ pos1,\ pos2,\ t1,\ t2)$,
which analogous can be seen as an degenerated three dimensional rectangle.

We use these NBB and TNBB to create Network Position Indexes (NPI) and
Temporal Network Position Indexes (TNPI) indexing the network positions of
the \dt{mgpoint} values of the \bmodb{}.

An NPI is an R-Tree build from the NBB
of the $route intervals$ of the $trajectory$ attributes of the \dt{mgpoint}
values.

An TNPI is an R-Tree build from the TNBB of the \dt{ugpoint} values of the
\dt{mgpoint} values.
\subsection{Network Operations used in \bmodb{}}
\label{sec:implNetOperations}
The operations used to construct the network object (\op{thenetwork}), and to
translate spatial and spatio-temporal values into network values are described in
Section \ref{sec:Translation}. In this section we give an overview of the
operations on network objects used in the benchmark queries of the \bmodb{}.

Table \ref{tab:simplenetoperations} gives an overview of the simple benchmark
operations. The more complex operations are described in the sequel.
\begin{sidewaystable}
  \caption{Simple Operations on Network Data Types}
  \label{tab:simplenetoperations}
  %\begin{scriptsize}
  \begin{center}
  \begin{tabularx}{1.0\textwidth}{|l|l|X|X|}
  \hline
  \textbf{Name}& \textbf{Signature}& \textbf{Explanation}& \textbf{Complexity}\\
  \hline
\op{routes}& \dt{network} $\rightarrow$ \dt{relation} & Returns the $routes$
relation fo the \dt{network} object & O($R$)\\
  \hline
\op{no\_components}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the number of
units of the \dt{mgpoint} &O(1)\\
  \hline
\op{length}&\dt{mgpoint} $\rightarrow$ \dt{real}& Returns the $length$ of the
argument.&O(1)\\
  \hline
\op{trajectory}&\dt{mgpoint} $\rightarrow$ \dt{gline}& Returns the $trajectory$
of the \dt{mgpoint} value as \dt{gline}&O($r$), if trajectory is defined, O($m \log r + r$) otherwise\\
  \hline
\op{units}&\dt{mgpoint} $\rightarrow$ \dt{stream}(\dt{ugpoint})& Returns a
stream
of the \dt{ugpoint} values of the \dt{mgpoint} value& O($m$)\\
  \hline
\op{initial}&\dt{mgpoint} $\rightarrow$ \dt{igpoint}& Returns the first position
and the start time of the \dt{mgpoint} value& O(1)\\
  \hline
\op{atinstant}&\dt{mgpoint} $\times$ \dt{instant} $\rightarrow$ \dt{igpoint}&
Returns the network position of the \dt{mgpoint} value at the given time instant&
O($\log m$)\\
  \hline
\op{val}&\dt{igpoint} $\rightarrow$ \dt{gpoint}& Returns the \dt{gpoint} of the
\dt{igpoint} value&O(1)\\
  \hline
\op{inst}&\dt{igpoint} $\rightarrow$ \dt{instant}& Returns the time instant of the
\dt{igpoint} value &O(1)\\
  \hline
\op{isempty}&\dt{gline} $\rightarrow$ \dt{bool}& Returns \true{} if the \dt{gline}
has no $route\ intervals$ & O(1)\\
  \hline
\op{routeintervals}&\dt{gline} $\rightarrow$ \dt{stream}(\dt{rectangle})& Returns
the $netbox$ for each $route\ interval$ of the \dt{gline} value& O($r$)\\
  \hline
\op{gpoint2rect}& \dt{gpoint} $\rightarrow$ \dt{rectangle} & Returns a $netbox$
for the \dt{gpoint} value & O(1)\\
\hline
\op{unitbox}& \dt{ugpoint} $\rightarrow$ \dt{rectangle3D} & Returns the $netbox$
of the \dt{ugpoint} & O(1)\\
  \hline
  \end{tabularx}
  \end{center}
  %\end{scriptsize}
\end{sidewaystable}

We use $m$ as the number of units of the \dt{mgpoint} value,
$r$ (resp. $r_1$, $r_2$) as the number of $route\ intervals$ of the \dt{gline}
value or the $trajectory$ attribute of an \dt{mgpoint} value (resp. of the
first, second argument), $R$ as the number of routes in the network object, $l_i$
the number of line segments of a route curve $r_i$,
and $p$ as the number of time intervals in a \dt{periods} value.

For the Euclidean Distance computation we retranslate our network values into
spatial (\op{gline2line}) respectively spatio-temporal (\op{mgpoint2mpoint})
values. In the original paper this operations are called \op{in\_space}.
\begin{tabbing}
xxxxxxxxxxxxxxxxxxxxxxxxxxxxxx\=xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx\kill
\dt{gline} $\rightarrow$ \dt{line} \> \op{gline2line}($gline$) \\
\dt{mgpoint} $\rightarrow$ \dt{mpoint} \> \op{mgpoint2mpoint}($mgpoint$)\\
\end{tabbing}
The operation \op{gline2line} uses the Algorithm \ref{alg:gline2line} to
translate a \dt{gline} value into an \dt{line} value.
\begin{algorithm}[H]
  \caption{\op{gline2line}($gl$)}
  \label{alg:gline2line}
  \begin{algorithmic}[1]
    \FOR {each $route\ interval$ of $gl$}
      \STATE Get route curve of $route\ interval$ using B-Tree index of $routes$ relation
      \STATE Perform binary search on line segments of route curve to find the
      start position of the $route\ interval$
      \STATE Copy line segments to result until end pos of $route\ interval$ is
reached.
    \ENDFOR
    \RETURN resulting line
  \end{algorithmic}
\end{algorithm}
The loop from line 1 to
line 5 is repeated $r$ times. The B-Tree search in line 2 take O($\log{R}$)
time. The binary search in line 3 takes O($\log l_i$) tim. The copy operation in line 4 takes
O($l_i$) time in the worst case. The return of the result has a worst case
time complexity of O($\sum_{i=1}^{r}{l_i}$). For the whole algorithm we
get
\[O(r \log{R} + \sum_{i=1}^{r}{\log{l_i}} + 2\sum_{i=1}^{r}{l_i})
= O(r \log{R} + \sum_{i=1}^{r}{l_i})\].
